MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-24 docs/paper/out/paper.tex
add BLAST ref, minor fixes in wording -- paper

\tiny
 \keyFont{ \section{Keywords:} Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
Cloud computing, NGS, Genomic big data } %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh} and
\href{https://github.com/ohnosequences/blast}{BLAST} expressions and
their execution.

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database. Assignment based on direct
similarity of each read one by one compared against a sufficiently wide
database is a very exhaustive method for assignment
\citep[\citet{morgan2012chapter}]{segata2013computational}. Some methods
of assignment compare the sequences only against the available complete
bacterial genomes or avoid computational cost clustering or binning the
sequences first, and then doing the assignments only for the
representative sequence of each cluster. MG7 carries out an exhaustive
comparison of all the reads under analysis and it does not applies any
binning strategy. Every read is specifically compared with all the
sequences of the 16S database. We select the best BLAST hits (10 hits by
default) obtained for each read to do the taxonomic assignment.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a BLAST expect value threshold (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest
E-value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below the product \(p S\) are filtered. p is a coefficient
fixed by the user to define the bitscore required, e.g.~if \(p=0.9\) and
\(S=700\) the required bitscore threshold would be \(630\). Once we have
the definitive HSPs selected, we obtain their corresponding taxonomic
nodes using the taxonomic assignments that NCBI provides for all the nt
database sequences. Now we have to analyze the topological distribution
of these nodes in the taxonomy tree: i. If all the nodes forms a line in
the taxonomy tree (are located in a not branched lineage to the tree
root) we should choose the most specific taxID as the final assignment
for that read. We call to this kind of assignment the `in line'
exception (see Figure 2 right panel). ii. If not, we should search for
the \emph{sensu stricto} Lowest Common Ancestor (LCA) of all the
selected taxonomic nodes (See Figure 2 left panel). In this approach we
decided to use the bitscore for evaluating the similarity because it is
a value that increases when similarity is higher and depends a lot on
the length of the HSP. Some reads could not find sequences with enough
similarity in the database and then they would be classified as reads
with no hits. Advanced metagenomics analysis approaches
\citep{huson2012microbial} have adopted LCA assignment algorithms
because it provides fine and trusted taxonomical assignment.

@@

MG7 provides independent results for the 2 different approaches, LCA and
BBH. The output files include, for each taxonomy node (with some read
assigned), abundance values for direct assignment and cumulative
assignment. The abundances are provided in counts (absolute values) and
in percentage normalized to the number of reads of each sample. Direct
assignments are calculated counting reads specifically assigned to a
taxonomic node, not including the reads assigned to the descendant nodes
in the taxonomy tree. Cumulative assignments are calculated including
the direct assignments and also the assignments of the descendant nodes.
For each sample MG7 provides 8 kinds of abundance values: LCA direct
counts, LCA cumu. counts, LCA direct \%, LCA cumu. \%, BBH direct
counts, BBH cumu. counts, BBH direct \%, BBH cumu. \%.

@@

The MG7 16 data analysis workflow is indeed a set of tasks, all of them
based in \emph{Loquat}. For each task, a set of inputs and outputs as
well as configuration parameters must be statically defined. The user is
also free to leave the reasonable defaults for configuration, needing
only to define the input and output of the whole workflow. The
definition of this configuration is Scala code and the way of starting
an MG7 analysis is compiling the project code and launching it from the
Scala interactive console.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treat data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  input and output data, their locations and type are expressible and
  checked at compile-time using \emph{Datasets}
\item
  management of dependencies and machine configurations using
  \emph{Statika}
\item
  automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

@@

\subsection{A new approach to data
analysis}\label{a-new-approach-to-data-analysis}

@@

\begin{itemize}
\tightlist
\item
  \textbf{Reproducibility} the same analysis can be run again with
  exactly the same configuration in a trivial way.
\item
  \textbf{Versioning} as in any software project, there can be different
  versions, stable releases, etc.
\item
  \textbf{Reuse} we can build standard configurations on top of this and
  reuse them for subsequent data analysis. A particular data analysis
  \emph{task} can be used as a \emph{library} in further analysis.
\item
  \textbf{Decoupling} We can start working on the analysis
  specification, without any need for available data in a much easier
  way.
\item
  \textbf{Documentation} We can take advantage of all the effort put
  into software documentation tools and practices, such as in our case
  Scaladoc or literate programming. As documentation, analysis processes
  and data specification live together in the files, it is much easier
  to keep coherence between them.
\item
  \textbf{Expresiveness and safety} For example in our case we can
  choose only from valid Illumina read types, and then build a default
  FLASh command based on that. The output locations, being declared
  statically, are also available for use in further analysis.
\end{itemize}

@@

An important aspect of the MG7 workflow is the way it deals with data
resources. All the data that is going to be used in the analysis or
produced as an output is described as Scala code using rich types from
the \emph{Datasets} language. This allows The user can specify
information about types of data, information that can then be utilized
by tools analyzing this data. For example, we can specify that for the
first part of the MG7 workflow running FLASh in parallel, requires
Illumina paired end reads and produces joined reads.

On one hand, specification of the input data allows us to restrict its
type and force users to be conscious about what they pass as an input.
On the other hand specification of the output data helps to build a
workflow as a \emph{composition} of several parts: we can ensure on the
Scala code type level that the output of one component fits as an input
for the next component. This is crucial, as obviously the way a data
analysis task works depends a lot on the particular structure of the
data. For instance, in the MG7 workflow, using BLAST eDSL, we can
precisely describe which format will have the output of the BLAST step,
which information it will include, and then in the next step we can
reuse this description to parse BLAST output and retrieve the part of
the information needed for the taxonomy assignment analysis. Having data
structure described statically as Scala code allows us to be sure that
we will not have parsing problems or other issues with incompatible data
passed between workflow components.

@@

Bioinformatics software often has a complicated installation process and
requires various dependencies with unclear versions. This makes the
deployment of the bioinformatics tools an involved task and resolving it
manually is not a solution in the context of cloud computations. To face
this problem, one needs an automated system of managing tools and
resources, which will allow an expressive way for describing
dependencies between parts of a pipeline and provide a reproducible
procedure of its deployment. We have developed \emph{Statika} for this
purpose and successfully use it in MG7.

Every external tool involved in the workflow is represented as a
\emph{Statika} bundle, which is essentially a Scala project describing
the installation process of this tool and declaring dependencies on
other bundles which will be installed prior to the considered tool
itself. Describing relationships between bundles on the code level
allows us to track the directed acyclic graph of their dependencies and
linearize them to automatically install them sequentially in the right
order. Meanwhile describing the installation process on the code level
allows the user to utilize the wide range of available Scala and Java
APIs and tools, making installation a well-defined sequence of steps
rather than an unreliable script, dependent on a certain environment.
\emph{Statika} offers an easy path towards making deployment an
automated, reproducible process.

@@

The MG7 workflow consists of certain steps, each of which performs some
work in parallel, using the cloud infrastructure managed by
\emph{Loquat}. It is important to notice the horizontal scalability of
this approach. Irrespectively of how much data needs to be processed,
MG7 will handle it, by splitting data into chunks and performing the
analysis on multiple computation units. The Amazon Elastic Compute Cloud
(EC2) service provides a transparent way of managing computation
infrastructure, called autoscaling groups. The User can set MG7
configuration parameters, adjusting for each task the amount and
hardware characteristics of the EC2 instances they want to use for it.
But it is important to note that, as each workflow step is not very
resource demanding, it is not needed to hire EC2 instances with some
advanced hardware, instead an average type will work and you can reduce
execution time by simply scaling out the number of instances.

@@

The hierarchic structure of the taxonomy of the living organisms is a
tree, and, hence, is also a graph in which each node, with the exception
of the root node, has a unique parent node. It led us to model the
taxonomy tree as a graph using the graph database paradigm. Previously
we developed Bio4j \textbf{{[}Pareja-Tobes-2015{]}}, a platform for the
integration of semantically rich biological data using typed graph
models. It integrates most publicly available data linked with sequences
into a set of interdependent graphs to be used for bioinformatics
analysis and especially for biological data.

@@

@@

>>>>

\tiny
 \keyFont{ \section{Keywords:} Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
Cloud computing, NGS, Genomic big data, Microbiome, Environmental, 16S
Database } %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh}
\citep{magovc2011flash} and
\href{https://github.com/ohnosequences/blast}{BLAST}
\citep{camacho2009blast} expressions and their execution.

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep[\citet{morgan2012chapter}]{segata2013computational} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database. We select the best BLAST hits (10 hits by default)
obtained for each read to do the taxonomic assignment.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity. To evaluate similarity for this first
filtering of hits we use the Expect value (by default
\(evalue \leq e^{-15}\)) that describes the number of hits one can
``expect'' to see by chance when searching a database of a particular
size. In a second filtering step we filtering those hits that are not
sufficiently good comparing them with the best one. We select the best
HSP (High Similarity Pair) per reference sequence and then choose the
best HSP (that with lowest E-value) between all the selected ones. The
bitscore of this best HSP (called S) is used as reference to filter the
rest of HSPs. All the HSPs with bitscore below the product \(p S\) are
filtered. p is a coefficient fixed by the user to define the bitscore
required, e.g.~if \(p=0.9\) and \(S=700\) the required bitscore
threshold would be \(630\). Once we have the definitive HSPs selected,
we obtain their corresponding taxonomic nodes using the taxonomic
assignments that NCBI provides for all the nt database sequences. Now we
have to analyze the topological distribution of these nodes in the
taxonomy tree: i. If all the nodes forms a line in the taxonomy tree
(are located in a not branched lineage to the tree root) we should
choose the most specific taxID as the final assignment for that read. We
call to this kind of assignment the `in line' exception (see Figure 2
right panel). ii. If not, we should search for the \emph{sensu stricto}
Lowest Common Ancestor (LCA) of all the selected taxonomic nodes (See
Figure 2 left panel). In this approach we decided to use the bitscore
for evaluating the similarity because it is a value that increases when
similarity is higher and depends a lot on the length of the HSP. Some
reads could not find sequences with enough similarity in the database
and then they would be classified as reads with no hits. Advanced
metagenomics analysis approaches \citep{huson2012microbial} have adopted
LCA assignment algorithms because it provides fine and trusted
taxonomical assignment.

@@

MG7 provides independent results for the 2 different approaches, LCA and
BBH. The output files include, for each taxonomy node (with some read
assigned), abundance values for direct assignment and cumulative
assignment. The abundances are provided in counts (absolute values) and
in percentage normalized to the number of reads of each sample. Direct
assignments are calculated counting reads specifically assigned to a
taxonomic node, not including the reads assigned to the descendant nodes
in the taxonomy tree. Cumulative assignments are calculated including
the direct assignments and also the assignments of the descendant nodes.
For each sample MG7 provides 8 kinds of abundance values: LCA direct
counts, LCA cumu. counts, LCA direct \%, LCA cumu. \%, BBH direct
counts, BBH cumu. counts, BBH direct \% and BBH cumu. \%.

@@

The MG7 16S data analysis workflow is indeed a set of tasks, all of them
based in \emph{Loquat}. For each task, a set of inputs and outputs as
well as configuration parameters must be statically defined. The user is
also free to leave the reasonable defaults for configuration, needing
only to define the input and output of the whole workflow. The
definition of this configuration is Scala code and the way of starting
an MG7 analysis is compiling the project code and launching it from the
Scala interactive console.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Treating data analysis as a software project. This makes for radical
  improvements in \emph{reproducibility}, \emph{reuse},
  \emph{versioning}, \emph{safety}, \emph{automation} and
  \emph{expressiveness}
\item
  Checking at compile-time: input and output data, their locations and
  type are expressible and checked at compile-time using \emph{Datasets}
\item
  Management of dependencies and machine configurations using
  \emph{Statika}
\item
  Automation of AWS cloud resources and processes, including
  distribution and parallelization through the use of \emph{Loquat}
\item
  Taxonomic data and related operations are treated natively as what
  they are: graphs, through the use of \emph{Bio4j}
\item
  MG7 provides a sustainable model for taxonomic assignment, appropriate
  to face the challenging amount of data that high throughput sequencing
  technologies generate
\end{enumerate}

@@

\subsection{A new approach to data analysis: data analysis as a software
project and checking at
compile-time}\label{a-new-approach-to-data-analysis-data-analysis-as-a-software-project-and-checking-at-compile-time}

@@

\begin{itemize}
\tightlist
\item
  \textbf{Reproducibility} the same analysis can be run again with
  exactly the same configuration in a trivial way.
\item
  \textbf{Versioning} as in any software project, there can be different
  versions, stable releases, etc.
\item
  \textbf{Reuse} we can build standard configurations on top of this and
  reuse them for subsequent data analysis. A particular data analysis
  \emph{task} can be used as a \emph{library} in further analysis.
\item
  \textbf{Decoupling} We can start working on the analysis
  specification, without any need for available data in a much easier
  way.
\item
  \textbf{Documentation} We can take advantage of all the effort put
  into software documentation tools and practices, such as in our case
  Scaladoc or literate programming. As documentation, analysis processes
  and data specification live together in the files, it is much easier
  to keep coherence between them.
\item
  \textbf{Expresiveness and safety} For example in our case we can
  choose only from valid illumina read types, and then build a default
  FLASh command based on that. The output locations, being declared
  statically, are also available for use in further analysis.
\end{itemize}

@@

An important aspect of the MG7 workflow is the way it deals with data
resources. All the data that is going to be used in the analysis or
produced as an output is described as Scala code using rich types from
the \emph{Datasets} language. This allows the user to specify
information about types of data, information that can then be utilized
by tools analyzing this data. For example, we can specify that, for the
first part of the MG7 workflow, running FLASh in parallel requires
illumina paired end reads and produces joined reads.

On one hand, specification of the input data allows us to restrict its
type and force users to be conscious about what they pass as an input.
On the other hand, specification of the output data helps to build a
workflow as a \emph{composition} of several parts: we can ensure on the
Scala code type level that the output of one component fits as an input
for the next component. This is crucial as, obviously, the way a data
analysis task works depends a lot on the particular structure of the
data. For instance, in the MG7 workflow, using BLAST eDSL, we can
precisely describe which format will have the output of the BLAST step,
which information it will include, and then in the next step we can
reuse this description to parse BLAST output and retrieve the part of
the information needed for the taxonomy assignment analysis. Having the
data structure described statically as Scala code allows us to be sure
that we will not have parsing problems or other issues with incompatible
data passed between workflow components.

@@

Bioinformatics software often has a complicated installation process and
requires various dependencies with unclear versions. This makes the
deployment of the bioinformatics tools an involved task and resolving it
manually is not a solution in the context of cloud computations. To face
this problem, one needs an automated system of managing tools and
resources, which will allow an expressive way for describing
dependencies between parts of a pipeline and provide a reproducible
procedure of its deployment. We have developed \emph{Statika} for this
purpose and successfully used it in MG7.

Every external tool involved in the workflow is represented as a
\emph{Statika} bundle, which is essentially a Scala project describing
the installation process of this tool and declaring dependencies on
other bundles which will be installed prior to the considered tool
itself. Describing relationships between bundles on the code level
allows us to track the directed acyclic graph of their dependencies and
linearize them to automatically install them sequentially in the right
order. Meanwhile, describing the installation process on the code level
allows the user to utilize the wide range of available Scala and Java
APIs and tools, making installation a well-defined sequence of steps
rather than an unreliable script, dependent on a certain environment.
\emph{Statika} offers an easy path towards making deployment an
automated, reproducible process.

@@

The MG7 workflow consists of certain steps, each of which performs some
work in parallel, using the cloud infrastructure managed by
\emph{Loquat}. It is important to notice the horizontal scalability of
this approach. Irrespectively of how much data needs to be processed,
MG7 will handle it, by splitting data into chunks and performing the
analysis on multiple computation units. The Amazon Elastic Compute Cloud
(EC2) service provides a transparent way of managing computation
infrastructure, called autoscaling groups. The User can set MG7
configuration parameters, adjusting for each task the amount and
hardware characteristics of the EC2 instances they want to use for it.
But it is important to note that, as each workflow step is not very
resource demanding, it is not needed to hire EC2 instances with some
advanced hardware. Instead, an average type will work and you can reduce
execution time by simply scaling out the number of instances.

@@

The hierarchic structure of the taxonomy of the living organisms is a
tree, and, hence, is also a graph in which each node, with the exception
of the root node, has a unique parent node. It led us to model the
taxonomy tree as a graph using the graph database paradigm. Previously
we developed Bio4j \textbf{{[}Pareja-Tobes-2015{]}}, a platform for the
integration of semantically rich biological data using typed graph
models. It integrates most publicly available data linked with sequences
into a set of interdependent graphs to be used for bioinformatics
analysis and especially for biological data. MG7 works based on the
Bio4j taxonomy module. It opens the possibility to connect the taxonomic
profiling data obtained with MG7 to all the biological knowledge
associated to each taxon. Using the information available in Bio4j for
all the proteins assigned to each taxon we are connected to all the
functional data available in Uniprot related with it.

@@

The comparison of the taxonomic profiles between different groups of
samples is a need for many metagenomics studies. Tasks related with this
group-based analysis, such as the extraction of the minimal tree with
all the taxa with some direct or accumulated assignment, will be part of
a new MG7 module, already in development.

@@

New visualization tools for metagenomics results are undoubtedly needed.
Interactivity is a especially interesting feature for metagenomics data
visualization, since the expert needs to explore the results in a
knowledge-driven way. The majority of the available metagenomics data
visualizations are static. We are working in the \emph{Biographika}
project \citep{tobes2015biographika}, to provide interactive rich
visualizations on the web for Bio4j data. The development of
visualizations specific for MG7 is one of Biographika current goals.
Biographika is based on D3.js, the de-facto standard JavaScript data
visualization library, and is open source.
